---
title: "Binomial Species Ratio Model"
author: "Sarah Popov"
date: '`r Sys.Date()`'
output: pdf_document
---

```{r binomial setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE)

library(ggplot2)

# This RMD is NOT intended as a standalone script, but rather to be sourced
# from models.Rmd. The data used in this Rmd file is generated in models.Rmd.
```

\newpage

# Binomial species ratio model

The daily ratio of Western sandpiper (WESA) to Dunlin (DUNL) across the entire study period is first modelled using a dataset of known species ratios (species ratios are not measured during every survey).

The ratios are modelled using a binomial generalized linear mixed model (binomial GLMM). The resulting predicted ratios are then used to estimate the number of WESA vs. DUNL per day.

## `sr` data summary

```{r sr summary}
summary(sr)
```

## Five models are built and compared

Response variable: 

* `y` - WESA:DUNL ratio

Predictor variables: 

* `dos` - day of season (recentered/scaled Julian date)
* `year` - year of survey

```{r sr glmm fitting}
# First, create our response variable: 
# WESA:DUNL proportion from our daily_percent_ratio table
y <- cbind(sr$wesa, sr$dunl)

# And build our 5 models 
# Response variable: ratio of WESA to DUNL
# Predictor variables:
# - dos: day of season (recentered/scaled julian date)
# - year: year (as a factor)
lme1 <- lme4::glmer(y ~ dos + I(dos^2) + (dos + I(dos^2)|year),
                    family = binomial, 
                    data = sr)
lme2 <- lme4::glmer(y  ~ dos + I(dos^2) + (1|year), 
                    family = binomial, 
                    data = sr)
lme3 <- lme4::glmer(y ~ dos + (dos|year), 
                    family = binomial, 
                    data = sr)
lme4 <- lme4::glmer(y ~ dos + (1|year), 
                    family = binomial, 
                    data = sr)
lme5 <- lme4::glmer(y ~ 1 + (1|year), 
                    family = binomial, 
                    data = sr)

## Compare AIC values 
anova(lme1, lme2, lme3, lme4, lme5)
```

The best-fit model is `lme1`. The residuals from `lme1` are appended to the `sr` dataset and another model is re-fit in order to estimate overdispersion. Because the standard deviation of the residuals is < 1, the model is deemed an appropriate candidate for predicting daily species ratios.

```{r best fit sr}
sr$resids <- residuals(lme1)

# Add residuals into model to examine coefficients
lme4::glmer(y ~ dos + I(dos^2) + (dos + I(dos^2)|year) + (1|resids),
            family = binomial, 
            data = sr)

sr_glmm <- lme1

# Add predicted values to sr
sr$predicted_ratio <- fitted(sr_glmm)

# Remove all the other lmes
rm(lme1, lme2, lme3, lme4, lme5)
```

\newpage
### Summary of best fit model (`lme1`)

```{r summary best fit sr}
summary(sr_glmm)
```

\newpage
### Check assumptions of best fit model


```{r sr plot: observed vs predicted}
ggplot(sr, aes(x = p_wesa, y = predicted_ratio)) +
  geom_point() + 
  ggtitle("Observed vs. Predicted values") +
  xlab("Observed") + 
  ylab("Predicted") +
  theme_minimal()
```


```{r sr plot: heteroskedasticity}
ggplot(sr, aes(x = predicted_ratio, y = resids)) +
  geom_point() + 
  ggtitle("Heteroskedasticity",
          subtitle = "Fitted values vs. Residuals") +
  xlab("Fitted") + 
  ylab("Residuals") +
  theme_minimal()
```


```{r sr plot: qq plot}
ggplot(sr, aes(sample = resids)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Quantile-Quantile") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```


```{r sr year effects, include = FALSE}
year_effects <- lme4::ranef(sr_glmm)$year
year_effects <- janitor::clean_names(year_effects)
```


```{r sr plot: year intercept plot}
ggplot(year_effects, aes(sample = intercept)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Year effects", 
          subtitle = "Intercept") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```


```{r sr plot: year slope plot}
ggplot(year_effects, aes(sample = dos)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Year effects", 
          subtitle = "Slope") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```


```{r sr plot: year slope quadratic plot}
ggplot(year_effects, aes(sample = i_dos_2)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Year effects", 
          subtitle = "Slope quadratic") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```

\newpage
## Predict WESA/DUNL population

Using the derived binomial GLMM above, we will predict the amount of WESA and DUNL each day. For any years that are missing from the bGLMM we will assign the mean proportion of WESA:DUNL. 

```{r predict wesa ratio, include = FALSE}
# Create yrs df, which is a df containing all possible
# julian_day-year combinations
yrs <- expand.grid(julian_day = seq(min(dt$julian_day), max(dt$julian_day), 1),
                   year = as.factor(unique(lubridate::year(dt$survey_date))))
# Add day of season variable
yrs$dos <- (yrs$julian_day - mean(sr$julian_day))/sd(sr$julian_day)

# Extract sr_glmm model coefficients
coefs <- coef(sr_glmm)$year
names(coefs) <- c("intercept", "dos_param", "dos2_param")
coefs$year <- rownames(coefs)

# Merge yrs and coefs together
yrs <- merge(yrs, coefs, by = "year", all.x = TRUE)

# Use mean coefficient values for any years where coefs NA
yrs[["intercept"]][is.na(yrs$intercept)] <- mean(coefs$intercept)
yrs[["dos_param"]][is.na(yrs$dos_param)] <- mean(coefs$dos_param)
yrs[["dos2_param"]][is.na(yrs$dos2_param)] <- mean(coefs$dos2_param)

# Now calculate odds & predict ratios for all survey dates
yrs$odds <- yrs$intercept + yrs$dos_param * yrs$dos + yrs$dos2_param * yrs$dos^2
yrs$predicted_ratio <- exp(yrs$odds) / (1 + exp(yrs$odds))
```

```{r add predicted ratios to dat, include=FALSE}
# Add predicted wesa ratio from binomial model to dat
dat <- merge(dat, yrs[,c("year", "julian_day", "predicted_ratio")], 
             by = c("year", "julian_day"))

# Scale and log-transform variables of interest
dat$predicted_wesa <- round(dat$final_count * dat$predicted_ratio, 0)
dat$predicted_dunl <- dat$final_count - dat$predicted_wesa
dat$log_wesa <- log(dat$predicted_wesa + 1)
dat$log_dunl <- log(dat$predicted_dunl + 1)
dat$year_c <- scale(as.numeric(dat$year))
```

```{r add predicted ratios to dt, include=FALSE}
# Now merge predicted wesa ratio generated above to the dt dataset
dt <- merge(dt, yrs[,c("year", "julian_day", "predicted_ratio")], 
            by = c("year", "julian_day"))

# Scale and log-transform variables of interest
dt$predicted_wesa <- round(dt$total_count * dt$predicted_ratio, 0)
dt$predicted_dunl <- dt$total_count - dt$predicted_wesa
dt$log_wesa <- log(dt$predicted_wesa + 1)
dt$log_dunl <- log(dt$predicted_dunl + 1)
dt$year_n <- as.numeric(dt$year)
dt$year_c <- scale(as.numeric(dt$year))
```


```{r plot predicted wesa ratios, fig.height = 8, fig.cap="Yearly calculated seasonal species ratios. In years where species ratio subsample surveys were not conducted (1991, 1992, 1994, 1995, 1998), the mean species ratio curve across all years was applied."}
ggplot() +
  geom_line(data = yrs, aes(x = julian_day, y = predicted_ratio, group = year)) +
  geom_point(data = sr, aes(x = julian_day, y = (p_wesa/100), group = year), color = 'blue') +
  facet_wrap(~ year, ncol = 3)
```

```{r cleanup, include = FALSE}
rm(yrs, coefs, year_effects, y)
```

