---
title: "`r params$spp` models"
author: "Sarah Popov"
date: "`r Sys.Date()`"
output: pdf_document
params:
  spp: "WESA"
  model_set: "ZINB"
---

```{r setup models, include=FALSE}
library(ggplot2)
library(formatR)

options(knitr.duplicate.label = "allow")

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE,
                      tidy.opts = list(width.cutoff = 60), 
                      tidy = TRUE)

# Color palette (using blue & yellow from wesanderson Zissou1 palette)
pal <- c(rgb(86, 152, 175, maxColorValue = 255), rgb(229, 205, 79, maxColorValue = 255))

# Source the data file, which connects to the bppeeps database and then queries and processes the data
# This will produce two datasets: 
# 1) dat [counts + environmental covariates]
# 2) sr [daily species ratios]
source(here::here("models", "prepare_dat.R"))
# Then calculate the binomial species ratio to get WESA vs DUNL counts
source(knitr::purl(here::here("models", "child_docs", "binomial_sr.Rmd")))
# Then summarize it by north/south station to get N S records
source(here::here("models", "prepare_dat_ns.R"))

# Choose species count data based on provided param
# Choose y var based on param
if (params$spp == "WESA") {
  dat_ns$y <- dat_ns$predicted_wesa
} else {
  dat_ns$y <- dat_ns$predicted_dunl
}

# Proportion of zeroes
dat_ns_p0 <- sum(dat_ns$y == 0) / length(dat_ns$y)
```

```{r build models, include = FALSE, eval=FALSE}
all_vars <- c("year_c", "dos", "n_s", "scale(flow)", "scale(mean_temp)", 
              "scale(elev_range)", "scale(total_precip)", "scale(u)", 
              "scale(windspd)", "tide", 
              "n_s:scale(flow)", "n_s:scale(mean_temp)", "n_s:scale(elev_range)",
              "n_s:scale(total_precip)", "n_s:scale(u)", "n_s:scale(windspd)")

all_models <- lapply(1:length(all_vars), function(n) data.frame(t(combn(all_vars, n))))
colnames(all_models[[1]]) <- "X1"
all_models <- dplyr::bind_rows(all_models)

# Now combine into model formula
# Also obliterating the rest of the dataframe bc we no longer need it
all_models <- apply(all_models, 1,
                    function(x) paste(x[!is.na(x)], collapse = " + "))

# Now remove models where interaction terms were added but the main effect is not there
# Logic: if 'n_s:scale(x)' is present, scale(x) should appear twice in in the formula.
# If 'n_s:scale(x)' is present but scale(x) is only present once, that means that just 
# the scale(x) in the interaction term is present.
ints <- c("flow", "mean_temp", "elev_range", "total_precip", "u", "windspd")

# Split into models with just main effects vs main + interactions
all_models_m <- all_models[!grepl(":", all_models)]
all_models_i <- all_models[grepl(":", all_models)]
length(all_models_m) + length(all_models_i) == length(all_models)

# Since all interaction terms are with n_s, drop any interaction models without 
# a standalone n_s term
all_models_i <- all_models_i[grepl("n_s +", all_models_i)]

# Remove records where main effect is not present but interaction term is
tmp <- list()
for (i in ints) {
  # Pull out all models with scale(x) AND n_s:scale(x) interaction
  # e.g., n_s + scale(flow) + n_s:scale(flow)
  x <- all_models_i[(grepl(paste0(" scale\\(", i), all_models_i) & grepl(paste0("n_s:scale\\(", i), all_models_i))]
  
  # Subset non-i interactions
  other_ints <- ints[!grepl(i, ints)]
  
  # Now remove any records where scale(x) and n_s:scale(x) AND n_s:scale(y) 
  # are present BUT scale(y) on it's own is not present
  # e.g., n_s + scale(flow) + n_s:scale(flow) + n_s:scale(mean_temp)
  for (j in other_ints) {
    x <- x[!(!grepl(paste0(" scale\\(", j), x) & grepl(paste0("n_s:scale\\(", j), x))]
  }
  
  tmp[[i]] <- x
}
rm(ints, other_ints, i, j, x)

tmp <- unlist(tmp)

all_models <- c(all_models_m, tmp)

rm(all_models_m, all_models_i, tmp, all_vars)

# Now build complete formula so that all models have a y term and a random intercept of year
all_models <- paste("y ~", all_models, "+ (1 | year)")
# all_models_gam <- gsub("scale", "s", all_models)
# all_models_gam <- gsub("(1 | year)", "s(year, bs = 're')", all_models_gam)
```


```{r def-assign-params, include=FALSE}
# See: https://stackoverflow.com/questions/49475303/rmarkdown-child-documents-do-not-detect-their-params
assign_params <- function(file) {
  text <- readLines(file)
  knit_params <- knitr::knit_params(text)
  params <<- purrr::map(knit_params, "value")
}

# See https://stackoverflow.com/questions/41979027/modify-params-in-r-notebooks
bindingIsLocked("params", env = .GlobalEnv) 
unlockBinding("params", env = .GlobalEnv)
```

# Data summary

Dataset: one count record per N/S region per survey date, `r nrow(dat_ns)` records. `r round(dat_ns_p0*100, 1)`% of the records are zeroes.

```{r station response distribution, fig.cap=paste("Histogram of", params$spp, "count per N/S group per survey date.")}
hist(dat_ns$y, breaks = 100, main = "", xlab = paste(params$spp, "count"))
```

```{r response vs all vars p1, fig.cap=paste("Count of", params$spp, "vs environmental covariates of interest.")}
facet_names <- c("dos" = "DOS", 
                 "elev_range" = "Tidal Amp.", 
                 "flow" = "Discharge", 
                 "mean_temp" = "Mean Temp.", 
                 "total_precip" = "Tot. Precip.", 
                 "u" = "u", 
                 "windspd" = "Windspd", 
                 "year" = "Year")

(dat_ns %>%
   dplyr::mutate(year = as.numeric(as.character(year))) %>%
   dplyr::select(y, year, dos, n_s, elev_range, flow, total_precip, mean_temp, u, windspd) %>%
   tidyr::pivot_longer(cols = -c(y, n_s), values_drop_na = TRUE, names_to = "var", values_to = "value") %>%
    ggplot(aes(x = value, y = y, color = n_s)) +
    geom_point(size = 0.3) +
    geom_smooth() +
    scale_x_continuous(n.breaks = 3) + 
    scale_y_continuous(labels = scales::label_comma()) +
    scale_color_manual(values = pal) +
    ggforce::facet_wrap_paginate(~ var, ncol = 3, nrow = 3, labeller = as_labeller(facet_names), scales = "free_x", page = 1) +
    coord_cartesian(ylim = c(0, NA)) +
    labs(x = "Value", 
         y = paste(params$spp, "count"),
       color = "Station") +
    theme_minimal())
```

\newpage
# Full model set

Models were selected with a suite of biologically relevant variables in mind. ZINB models are zero-inflated negative binomial models fitted using the `glmmTMB` package. All zero-inflated models contain the same terms for both the conditional and zero-inflation component models.

'Full' models include all of the following predictors (continuous variables are all centered on their means and scaled by their standard deviations): 

* $DOS + DOS^2$, or quadratic effect of day of season
* $NS$, or North/South section
* $FLOW$, or river discharge
* $AMP$, or tidal amplitude
* $TEMP$, or mean temp
* $PRECIP$, or total precipitation
* $U$, or westerly wind


```{r LME model set, include=FALSE, eval=params$model_set == "LME"}
######################## BASE MODELS ########################
# A series of nested linear models
library(lme4)

base0 <- lm(formula("y ~ n_s + dos + I(dos^2)"),
              data = dat_ns, 
              control = lmerControl(optimizer="bobyqa"), 
              REML = FALSE) # fails to converge

base1 <- lmer(formula("y ~ n_s + dos + I(dos^2) + (1 | year)"), # different RE
              data = dat_ns, 
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

base2 <- lmer(formula("y ~ n_s + year_c + dos + I(dos^2) + (1 | year)"), # add year as predictor
              data = dat_ns, 
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

base3 <- lmer(formula("y ~ n_s * year_c + dos + I(dos^2) + (1 | year)"), # interaction between n_s and year
              data = dat_ns,
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

base4 <- lmer(formula("y ~ n_s + scale(flow) + I(dos^2) + (1 | year)"), # replace year with flow
              data = dat_ns,
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

base5 <- lmer(formula("y ~ n_s * scale(flow) + I(dos^2) + (1 | year)"), # interaction between n_s and flow
              data = dat_ns,
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

baseList <- list('NS + DOS + DOS^2' = base0,
                 'NS + DOS + DOS^2 + Year Intercept' = base1,
                 'NS + YEAR + DOS + DOS^2 + Year Intercept' = base2,
                 'NS * YEAR + DOS + DOS^2 + Year Intercept' = base3,
                 'NS + FLOW + DOS^2 + Year Intercept' = base4,
                 'NS * FLOW + DOS^2 + Year Intercept' = base5
                 )
```

```{r, ZINB model set, include=FALSE, eval=params$model_set == "ZINB"}
######################## ZINB MODELS ########################
# A series of nested zero-inflated negative binomial models
library(glmmTMB)

zinb0 <- glmmTMB(y ~ n_s + year_c + dos + I(dos^2) + (1|year), 
                 zi = ~ ., 
                 data = dat_ns, 
                 family = nbinom1, 
                 REML = FALSE)

zinb1 <- glmmTMB(y ~ n_s + scale(flow) + year_c + dos + I(dos^2) + (1|year), 
                 zi = ~ ., 
                 data = dat_ns, 
                 family = nbinom1,
                 REML = FALSE)

zinb2 <- glmmTMB(y ~ n_s * scale(flow) + year_c + dos + I(dos^2) + (1|year), 
                 zi = ~ ., 
                 data = dat_ns, 
                 family = nbinom1,
                 REML = FALSE)

zinbf0 <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide + dos + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf1 <- glmmTMB(y ~ n_s * scale(flow) + year_c 
                     + n_s * scale(mean_temp) 
                     + n_s * scale(elev_range) 
                     + n_s * scale(total_precip)
                     + n_s * scale(u) 
                     + n_s * scale(windspd) 
                     + tide + dos + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_flow <- glmmTMB(y ~ n_s * scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide + dos + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_temp <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + n_s * scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide + dos + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_ta <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + n_s * scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide + dos + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_tp <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + n_s * scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide + dos + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_u <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + n_s * scale(u) 
                     + scale(windspd) 
                     + tide + dos + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_ws <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + n_s * scale(windspd) 
                     + tide + dos + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbList <- list('ZINB + NS + YEAR + DOS + DOS^2 + Year Intercept' = zinb0,
                 'ZINB + NS + FLOW + YEAR + DOS^2 + Year Intercept' = zinb1,
                 'ZINB + NS * FLOW + YEAR + DOS^2 + Year Intercept' = zinb2,
                 'Full ZINB + Year Intercept' = zinbf0,
                 'Full ZINB + All interaction terms + Year Intercept' = zinbf1,
                 'Full ZINB + NS:FLOW + Year Intercept' = zinbf0_flow,
                 'Full ZINB + NS:TIDE_AMP + Year Intercept' = zinbf0_ta,
                 'Full ZINB + NS:TEMP + Year Intercept' = zinbf0_temp,
                 'Full ZINB + NS:PRECIP + Year Intercept' = zinbf0_tp,
                 'Full ZINB + NS:U + Year Intercept' = zinbf0_u,
                 'Full ZINB + NS:WINDSPD + Year Intercept' = zinbf0_ws)
```

```{r tgam model set, include=FALSE, eval=params$model_set == "GAM"}
######################## TWEEDIE GAM MODELS ########################
# A series of tweedie distribution general additive models
library(mgcv)

tgam0 <- gam(y ~ n_s + s(year_c) + s(dos) + s(year, bs = "re"), 
             data = dat_ns, 
             family = tw, 
             method = "ML")

tgam1 <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos) + s(year, bs = "re"), 
             data = dat_ns, 
             family = tw, 
             method = "ML")

tgam2 <- gam(y ~ n_s + s(flow, by = n_s) + s(year_c) + s(dos) + s(year, bs = "re"), 
             data = dat_ns, 
             family = tw, 
             method = "ML")

tgamf0 <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos) 
              + s(mean_temp) + s(elev_range) + s(total_precip)
              + s(u) + s(windspd) + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf1 <- gam(y ~ n_s + s(flow) + s(flow, by = n_s) + s(year_c) + s(dos)
              + s(mean_temp) + s(mean_temp, by = n_s) 
              + s(elev_range) + s(elev_range, by = n_s) 
              + s(total_precip) + s(total_precip, by = n_s) 
              + s(u) + s(u, by = n_s) 
              + s(windspd) + s(windspd, by = n_s) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_flow <- gam(y ~ n_s + s(flow) + s(flow, by = n_s) + s(year_c) + s(dos)
              + s(mean_temp)
              + s(elev_range) 
              + s(total_precip)
              + s(u)
              + s(windspd)
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_temp <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp) + s(mean_temp, by = n_s) 
              + s(elev_range) 
              + s(total_precip)
              + s(u)
              + s(windspd) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_ta <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp) 
              + s(elev_range) + s(elev_range, by = n_s) 
              + s(total_precip) 
              + s(u) 
              + s(windspd) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_tp <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp)
              + s(elev_range)
              + s(total_precip) + s(total_precip, by = n_s) 
              + s(u)
              + s(windspd)
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_u <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp)
              + s(elev_range) 
              + s(total_precip) 
              + s(u) + s(u, by = n_s) 
              + s(windspd) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_ws <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp)
              + s(elev_range) 
              + s(total_precip)
              + s(u)
              + s(windspd) + s(windspd, by = n_s) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamList <- list('tGAM + NS + YEAR + s(DOS) + Year Intercept' = tgam0,
                 'tGAM + NS + FLOW + YEAR + s(DOS) + Year Intercept' = tgam1,
                 'tGAM + NS * FLOW + YEAR + s(DOS) + Year Intercept' = tgam2,
                 'Full tGAM + Year Intercept' = tgamf0,
                 'Full tGAM + All interaction terms + s(DOS) + Year Intercept' = tgamf1,
                 'Full tGAM + NS:FLOW + s(DOS) + Year Intercept' = tgamf0_flow,
                 'Full tGAM + NS:TIDE_AMP + s(DOS) + Year Intercept' = tgamf0_ta,
                 'Full tGAM + NS:TEMP + s(DOS) + Year Intercept' = tgamf0_temp,
                 'Full tGAM + NS:PRECIP + s(DOS) + Year Intercept' = tgamf0_tp,
                 'Full tGAM + NS:U + s(DOS) + Year Intercept' = tgamf0_u,
                 'Full tGAM + NS:WINDSPD + s(DOS) + Year Intercept' = tgamf0_ws)
```

```{r AIC tab, include=FALSE}
modList <- c(baseList, zinbList, tgamList)

tab <- bbmle::AICtab(modList, weights = TRUE)
```

```{r AICtab, fig.cap="A comparison of all 28 models."}
knitr::kable(tab, digits = 3)
```

# Multi-model inference

Typically, the top $N$ number of candidate models that add up to a weight of 0.95 are chosen as the 'candidate set' of models and averaged together for multi-model inference (i.e., models with a 95% cumulative probability of containing the AIC-best model should the data be independently recollected). In this case, however, the top model (full tGAM with all covariates and all interaction terms and random year intercept) carries effectively all the AIC weight. 

In this case what I have instead done is average together the top 5 non-full models from the ZINB and GAM models to compare to the full models.

```{r extract multi model avg, include=FALSE}
tab2 <- as.data.frame(tab)
tab2$mod <- row.names(tab2)
rownames(tab2) <- seq(nrow(tab2))

# Cut out FULL models with EVERYTHING
tab2 <- tab2[!(tab2$mod %in% c("Full tGAM + All interaction terms + s(DOS) + Year Intercept", "Full ZINB + All interaction terms + Year Intercept")),]

tab2zinb <- tab2[grep("ZINB", tab2$mod),]
tab2tgam <- tab2[grep("tGAM", tab2$mod),]

options(scipen = 50) # turn of sci notation for small numbers

# Recalculate dAIC and weight
recalc_ICtab <- function(x) {
  x$dAIC <- x$dAIC - min(x$dAIC)
  x$evidence_ratio <- exp(-0.5 * (x$dAIC))
  x$weight <- x$evidence_ratio / sum(x$evidence_ratio)
  x$weight <- format.pval(x$weight,eps=0.001, digits=5)
  x <- x[,c("mod", "dAIC", "df", "weight")]
  rownames(x) <- seq(nrow(x))
  return(x)
}

tab2zinb <- recalc_ICtab(tab2zinb)
tab2tgam <- recalc_ICtab(tab2tgam)

top5mods <- function(x) {
  top5 <- x[1:5, c("mod")]
  y <- lapply(top5, function(z) modList[[z]])
  return(y)
}
```

## The top 5 ZINB models

The table below pulls out the top 5 ZINB models, *excluding* the model with all interaction terms. In other words, it ranks the models by the NS interaction terms that bring the candidate model the closest to the 'true' model.

```{r zinb top5 table}
knitr::kable(tab2zinb)
```

The top 5 ZINB models were then averaged together and presented below. 

```{r zinb top 5 model avg plot, fig.cap="Plot of averaged model coefficients from the top-5 ZINB models (excluding the full ZINB model with all coefficients and interaction terms)."}
sjPlot::plot_model(MuMIn::model.avg(top5mods(tab2zinb)))
```

```{r zinb top 5 model avg summary, fig.cap="Model summary of the averaged top-5 ZINB models (excluding the full ZINB model with all coefficients and interaction terms).", include=FALSE}
summary(MuMIn::model.avg(top5mods(tab2zinb)))
```

## The top 5 Tweedie GAMs

The table below pulls out the top 5 tweedie GAMs, *excluding* the model with all interaction terms. In other words, it ranks the models by the NS interaction terms that bring the candidate model the closest to the 'true' model.

Because the interpretation of averaged GAMs is unclear to me (it is not clear if/how the smoother terms are averaged), I have not included any 'averaged' GAM results.

```{r tgam top5 table}
knitr::kable(tab2tgam)
```

\newpage
# Full Tweedie GAM

Summaries of the full models with all interaction terms are presented below for comparison.

## Model summary

```{r summary tgam}
summary(tgamf1)
```

\newpage
```{r tgam estimated smooths, fig.cap="Fitted smooths from the full Tweedie GAM. Interaction term smoothers are plotted by N first, then by S (it gets cut off).", fig.height = 8, fig.width=6}
gratia::draw(tgamf1)
```

\newpage
```{r tgam diagnostics, fig.cap="Full Tweedie GAM diagnostics."}
gratia::appraise(tgamf1)
```

\newpage
# Full ZINB model

## Model summary

```{r summary zinbf1}
summary(zinbf1)
```

\newpage
```{r zinb coefficients, fig.cap="Standardized coefficient effect sizes from the full ZINB model."}
dat_ns$dos <- dat_ns$dos[,1]
dat_ns$year_c <- dat_ns$year_c[,1]
zinbf1 <- update(zinbf1, data = dat_ns)
sjPlot::plot_model(zinbf1, type = "std")
```

\newpage
```{r zinb diagnostics, fig.cap="Full ZINB model diagnostics.", echo=FALSE, results='hide', fig.keep='all'}
DHARMa::testResiduals(zinbf1)
```



