---
title: "`r params$spp` final models"
author: "Sarah Popov"
date: "`r Sys.Date()`"
output: pdf_document
params:
  spp: "WESA"
---

```{r setup models, include=FALSE}
library(ggplot2)
library(formatR)

options(knitr.duplicate.label = "allow")

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE,
                      tidy.opts = list(width.cutoff = 60), 
                      tidy = TRUE)

# Color palette (using blue & yellow from wesanderson Zissou1 palette)
pal <- c(rgb(86, 152, 175, maxColorValue = 255), rgb(229, 205, 79, maxColorValue = 255))

# Source the data file, which connects to the bppeeps database and then queries and processes the data
# This will produce two datasets: 
# 1) dat [counts + environmental covariates]
# 2) sr [daily species ratios]
source(here::here("models", "prepare_dat.R"))
# Then calculate the binomial species ratio to get WESA vs DUNL counts
source(knitr::purl(here::here("models", "child_docs", "binomial_sr.Rmd")))
# Then summarize it by north/south station to get N S records
source(here::here("models", "prepare_dat_ns.R"))

# Choose species count data based on provided param
# Choose y var based on param
if (params$spp == "WESA") {
  dat_ns$y <- dat_ns$predicted_wesa
} else {
  dat_ns$y <- dat_ns$predicted_dunl
}

# Proportion of zeroes
dat_ns_p0 <- sum(dat_ns$y == 0) / length(dat_ns$y)

```

```{r def-assign-params, include=FALSE}
# See: https://stackoverflow.com/questions/49475303/rmarkdown-child-documents-do-not-detect-their-params
assign_params <- function(file) {
  text <- readLines(file)
  knit_params <- knitr::knit_params(text)
  params <<- purrr::map(knit_params, "value")
}

# See https://stackoverflow.com/questions/41979027/modify-params-in-r-notebooks
bindingIsLocked("params", env = .GlobalEnv)
unlockBinding("params", env = .GlobalEnv)
```

# Data summary

Dataset: one count record per N/S region per survey date, `r nrow(dat_ns)` records. `r round(dat_ns_p0*100, 1)`% of the records are zeroes.

```{r station response distribution, fig.cap=paste("Histogram of", params$spp, "count per N/S group per survey date.")}
hist(dat_ns$y,  breaks = 100, main = "Frequency", xlab = paste(params$spp, "count"))
```

```{r response vs all vars p1}
facet_names <- c("dos" = "DOS", 
                 "elev_range" = "Tidal Amp.", 
                 "flow" = "Discharge", 
                 "mean_temp" = "Mean Temp.", 
                 "total_precip" = "Tot. Precip.", 
                 "u" = "u", 
                 "windspd" = "Windspd", 
                 "year" = "Year")

(dat_ns %>%
   dplyr::mutate(year = as.numeric(as.character(year))) %>%
   dplyr::select(y, year, dos, n_s, elev_range, flow, total_precip, mean_temp, u, windspd) %>%
   tidyr::pivot_longer(cols = -c(y, n_s), values_drop_na = TRUE, names_to = "var", values_to = "value") %>%
    ggplot(aes(x = value, y = y, color = n_s)) +
    geom_point(size = 0.3) +
    geom_smooth() +
    scale_x_continuous(n.breaks = 3) + 
    scale_color_manual(values = pal) +
    ggforce::facet_wrap_paginate(~ var, ncol = 3, nrow = 3, labeller = as_labeller(facet_names), scales = "free_x", page = 1) +
    coord_cartesian(ylim = c(0, 200000)) +
    labs(x = "Value", 
         y = paste(params$spp, "count"),
       color = "Station") +
    theme_minimal())
```

# Final models

Candidate models selected from a suite of biologically relevant models.

'Full' models include all of the following predictors: 
    - `FLOW`, or river discharge
    - `TIDE_AMP`, or tidal amplitude
    - `TEMP`, or mean temp
    - `PRECIP`, or total precipitation
    - `U`, or westerly wind
    - `WINDSPD`, or wind speed

```{r full model set, include=FALSE}
######################## BASE MODELS ########################
# A series of nested linear models
library(lme4)

base0 <- lmer(formula("y ~ n_s + dos + I(dos^2)"),
              data = dat_ns, 
              control = lmerControl(optimizer="bobyqa"), 
              REML = FALSE) # fails to converge

base1 <- lmer(formula("y ~ n_s + dos + I(dos^2) + (1 | year)"), # different RE
              data = dat_ns, 
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

base2 <- lmer(formula("y ~ n_s + year_c + dos + I(dos^2) + (1 | year)"), # add year as predictor
              data = dat_ns, 
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

base3 <- lmer(formula("y ~ n_s * year_c + dos + I(dos^2) + (1 | year)"), # interaction between n_s and year
              data = dat_ns,
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

base4 <- lmer(formula("y ~ n_s + scale(flow) + I(dos^2) + (1 | year)"), # replace year with flow
              data = dat_ns,
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

base5 <- lmer(formula("y ~ n_s * scale(flow) + I(dos^2) + (1 | year)"), # interaction between n_s and flow
              data = dat_ns,
              control = lmerControl(optimizer="bobyqa"),
              REML = FALSE)

# baseList <- list('NS + DOS + DOS^2' = base0,
#                  'NS + DOS + DOS^2 + Year Intercept' = base1,
#                  'NS + YEAR + DOS + DOS^2 + Year Intercept' = base2,
#                  'NS * YEAR + DOS + DOS^2 + Year Intercept' = base3,
#                  'NS + FLOW + DOS^2 + Year Intercept' = base4,
#                  'NS * FLOW + DOS^2 + Year Intercept' = base5
#                  )

######################## ZINB MODELS ########################
# A series of nested zero-inflated negative binomial models
library(glmmTMB)

zinb0 <- glmmTMB(y ~ n_s + year_c + dos + I(dos^2) + (1|year), 
                 zi = ~ ., 
                 data = dat_ns, 
                 family = nbinom1, 
                 REML = FALSE)

zinb1 <- glmmTMB(y ~ n_s + scale(flow) + year_c + I(dos^2) + (1|year), 
                 zi = ~ ., 
                 data = dat_ns, 
                 family = nbinom1,
                 REML = FALSE)

zinb2 <- glmmTMB(y ~ n_s * scale(flow) + year_c + I(dos^2) + (1|year), 
                 zi = ~ ., 
                 data = dat_ns, 
                 family = nbinom1,
                 REML = FALSE)

zinbf0 <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide  + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf1 <- glmmTMB(y ~ n_s * scale(flow) + year_c 
                     + n_s * scale(mean_temp) 
                     + n_s * scale(elev_range) 
                     + n_s * scale(total_precip)
                     + n_s * scale(u) 
                     + n_s * scale(windspd) 
                     + tide  + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_flow <- glmmTMB(y ~ n_s * scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide  + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_temp <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + n_s * scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide  + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_ta <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + n_s * scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide  + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_tp <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + n_s * scale(total_precip)
                     + scale(u) 
                     + scale(windspd) 
                     + tide  + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_u <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + n_s * scale(u) 
                     + scale(windspd) 
                     + tide  + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

zinbf0_ws <- glmmTMB(y ~ n_s + scale(flow) + year_c 
                     + scale(mean_temp) 
                     + scale(elev_range) 
                     + scale(total_precip)
                     + scale(u) 
                     + n_s * scale(windspd) 
                     + tide  + I(dos^2) + (1|year), 
                     zi = ~ ., 
                     data = dat_ns, 
                     family = nbinom1,
                     REML = FALSE)

# zinbList <- list('NS + YEAR + DOS + DOS^2 + Year Intercept + ZI' = zinb0,
#                  'NS + FLOW + YEAR + DOS^2 + Year Intercept + ZI' = zinb1,
#                  'NS * FLOW + YEAR + DOS^2 + Year Intercept + ZI' = zinb2,
#                  'Full ZINB + Year Intercept + ZI' = zinbf0,
#                  'Full ZINB + All interaction terms + Year Intercept + ZI' = zinbf1,
#                  'Full ZINB + NS:FLOW + Year Intercept + ZI' = zinbf0_flow,
#                  'Full ZINB + NS:TIDE_AMP + Year Intercept + ZI' = zinbf0_ta,
#                  'Full ZINB + NS:TEMP + Year Intercept + ZI' = zinbf0_temp,
#                  'Full ZINB + NS:PRECIP + Year Intercept + ZI' = zinbf0_tp,
#                  'Full ZINB + NS:U + Year Intercept + ZI' = zinbf0_u,
#                  'Full ZINB + NS:WINDSPD + Year Intercept + ZI' = zinbf0_ws)

######################## TWEEDIE GAM MODELS ########################
# A series of tweedie distribution general additive models
library(mgcv)

tgam0 <- gam(y ~ n_s + s(year_c) + s(dos) + s(year, bs = "re"), 
             data = dat_ns, 
             family = tw, 
             method = "ML")

tgam1 <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos) + s(year, bs = "re"), 
             data = dat_ns, 
             family = tw, 
             method = "ML")

tgam2 <- gam(y ~ n_s + s(flow, by = n_s) + s(year_c) + s(dos) + s(year, bs = "re"), 
             data = dat_ns, 
             family = tw, 
             method = "ML")

tgamf0 <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos) 
              + s(mean_temp) + s(elev_range) + s(total_precip)
              + s(u) + s(windspd) + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf1 <- gam(y ~ n_s + s(flow) + s(flow, by = n_s) + s(year_c) + s(dos)
              + s(mean_temp) + s(mean_temp, by = n_s) 
              + s(elev_range) + s(elev_range, by = n_s) 
              + s(total_precip) + s(total_precip, by = n_s) 
              + s(u) + s(u, by = n_s) 
              + s(windspd) + s(windspd, by = n_s) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_flow <- gam(y ~ n_s + s(flow) + s(flow, by = n_s) + s(year_c) + s(dos)
              + s(mean_temp)
              + s(elev_range) 
              + s(total_precip)
              + s(u)
              + s(windspd)
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_temp <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp) + s(mean_temp, by = n_s) 
              + s(elev_range) 
              + s(total_precip)
              + s(u)
              + s(windspd) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_ta <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp) 
              + s(elev_range) + s(elev_range, by = n_s) 
              + s(total_precip) 
              + s(u) 
              + s(windspd) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_tp <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp)
              + s(elev_range)
              + s(total_precip) + s(total_precip, by = n_s) 
              + s(u)
              + s(windspd)
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_u <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp)
              + s(elev_range) 
              + s(total_precip) 
              + s(u) + s(u, by = n_s) 
              + s(windspd) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

tgamf0_ws <- gam(y ~ n_s + s(flow) + s(year_c) + s(dos)
              + s(mean_temp)
              + s(elev_range) 
              + s(total_precip)
              + s(u)
              + s(windspd) + s(windspd, by = n_s) 
              + tide + s(year, bs = "re"), 
              data = dat_ns, 
              family = tw, 
              method = "ML")

# tgamList <- list('NS + YEAR + s(DOS) + Year Intercept' = tgam0,
#                  'NS + FLOW + YEAR + s(DOS) + Year Intercept' = tgam1,
#                  'NS * FLOW + YEAR + s(DOS) + Year Intercept' = tgam2,
#                  'Full tGAM + Year Intercept' = tgamf0,
#                  'Full tGAM + All interaction terms + s(DOS) + Year Intercept' = tgamf1,
#                  'Full tGAM + NS:FLOW + s(DOS) + Year Intercept' = tgamf0_flow,
#                  'Full tGAM + NS:TIDE_AMP + s(DOS) + Year Intercept' = tgamf0_ta,
#                  'Full tGAM + NS:TEMP + s(DOS) + Year Intercept' = tgamf0_temp,
#                  'Full tGAM + NS:PRECIP + s(DOS) + Year Intercept' = tgamf0_tp,
#                  'Full tGAM + NS:U + s(DOS) + Year Intercept' = tgamf0_u,
#                  'Full tGAM + NS:WINDSPD + s(DOS) + Year Intercept' = tgamf0_ws)

modList <- list('NS + DOS + DOS^2' = base0,
                 'NS + DOS + DOS^2 + Year Intercept' = base1,
                 'NS + YEAR + DOS + DOS^2 + Year Intercept' = base2,
                 'NS * YEAR + DOS + DOS^2 + Year Intercept' = base3,
                 'NS + FLOW + DOS^2 + Year Intercept' = base4,
                 'NS * FLOW + DOS^2 + Year Intercept' = base5,
                 'NS + YEAR + DOS + DOS^2 + Year Intercept + ZI' = zinb0,
                 'NS + FLOW + YEAR + DOS^2 + Year Intercept + ZI' = zinb1,
                 'NS * FLOW + YEAR + DOS^2 + Year Intercept + ZI' = zinb2,
                 'Full ZINB + Year Intercept + ZI' = zinbf0,
                 'Full ZINB + All interaction terms + Year Intercept + ZI' = zinbf1,
                 'Full ZINB + NS:FLOW + Year Intercept + ZI' = zinbf0_flow,
                 'Full ZINB + NS:TIDE_AMP + Year Intercept + ZI' = zinbf0_ta,
                 'Full ZINB + NS:TEMP + Year Intercept + ZI' = zinbf0_temp,
                 'Full ZINB + NS:PRECIP + Year Intercept + ZI' = zinbf0_tp,
                 'Full ZINB + NS:U + Year Intercept + ZI' = zinbf0_u,
                 'Full ZINB + NS:WINDSPD + Year Intercept + ZI' = zinbf0_ws,
                 'NS + YEAR + s(DOS) + Year Intercept' = tgam0,
                 'NS + FLOW + YEAR + s(DOS) + Year Intercept' = tgam1,
                 'NS * FLOW + YEAR + s(DOS) + Year Intercept' = tgam2,
                 'Full tGAM + Year Intercept' = tgamf0,
                 'Full tGAM + All interaction terms + s(DOS) + Year Intercept' = tgamf1,
                 'Full tGAM + NS:FLOW + s(DOS) + Year Intercept' = tgamf0_flow,
                 'Full tGAM + NS:TIDE_AMP + s(DOS) + Year Intercept' = tgamf0_ta,
                 'Full tGAM + NS:TEMP + s(DOS) + Year Intercept' = tgamf0_temp,
                 'Full tGAM + NS:PRECIP + s(DOS) + Year Intercept' = tgamf0_tp,
                 'Full tGAM + NS:U + s(DOS) + Year Intercept' = tgamf0_u,
                 'Full tGAM + NS:WINDSPD + s(DOS) + Year Intercept' = tgamf0_ws
                )
tab <- bbmle::AICtab(modList, base = TRUE, weights = TRUE)

# tab <-bbmle::AICtab(base0, base1, base2, base3, base4, base5, # base linear models
#               zinb0, zinb1, zinb2, # base zinb models
#               zinbf0, zinbf1, # full zinb models - 0 = no interactions, 1 = all interactions
#               zinbf0_flow, zinbf0_temp, zinbf0_ta, zinbf0_tp, zinbf0_u, zinbf0_ws, # zinb models with one interaction each
#               tgam0, tgam1, tgam2, # base tweedie gam models
#               tgamf0, tgamf1, # full tgam models - 0 = no interactions, 1 = all interactions
#               tgamf0_flow, tgamf0_temp, tgamf0_ta, tgamf0_tp, tgamf0_u, tgamf0_ws,
#               base = TRUE,
#               weights = TRUE
#               )
```




