---
title: "Mechanism (flow) vs time (year) models"
author: "Sarah Popov"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r mech vs time setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE)

# This RMD is NOT intended as a standalone script, but rather to be sourced
# from models.Rmd. The data used in this Rmd file is generated in models.Rmd
# and binomial_sr.Rmd.
```

# Exploring mechanism (flow) vs. time (year)

Given the Canham extended model has some issues, the next step is to run two models with a slightly simpler approach. We are still interested in seeing if the discharge (`flow`) has an impact on the north-south distribution of peeps. In addition, we are interested in seeing any temporal changes in this distribution.

Two simpler models can get at these questions:

1. The 'mechanism' model: `log_wesa ~ n_s * scale(flow) + dos + I(dos^2) + (dos + I(dos^2) | year)`
2. The 'temporal' model: `log_wesa ~ n_s * year_c + dos + I(dos^2) + (dos + I(dos^2) | year)`

## Data changes

These simpler models necessitate further data aggregation. We are primarily interested in N vs S distribution of peeps: therefore, the data are grouped by N/S such that there are two data points per survey date: number of peeps in the northern stations (Canoe Pass, Brunswick Point, Bend, Pilings, View corner) vs southern (34th St pullout, Coal Port). Any NA values in our variables of interest are also removed.

```{r aggregate by NS, include=FALSE}
dat3 <- sqldf::sqldf("select year, 
                      survey_date, 
                      julian_day, 
                      min(start_time) as start_time, 
                      n_s, 
                      sum(final_count) as final_count, 
                      sum(predicted_wesa) as predicted_wesa, 
                      sum(predicted_dunl) as predicted_dunl, 
                      p_wesa, 
                      predicted_ratio, 
                      avg(raptor_count) as raptor_count, 
                      elev_min, 
                      elev_max, 
                      elev_median, 
                      elev_mean, 
                      elev_range, 
                      flow, 
                      total_precip, 
                      mean_temp, 
                      u, 
                      v, 
                      windspd, 
                      wind_deg 
                      from dat 
                      group by survey_date, n_s;") %>%
  dplyr::mutate(dos = scale(julian_day),
                log_wesa = log(predicted_wesa + 1),
                log_dunl = log(predicted_dunl + 1),
                year_c = scale(as.numeric(year))) %>%
  dplyr::select(year, survey_date, julian_day, dos, start_time, n_s,
                final_count, predicted_wesa, predicted_dunl, log_wesa, 
                log_dunl, dplyr::everything()) %>%
  dplyr::filter(!is.na(flow), !is.na(total_precip))
```

```{r head dat3}
head(dat3)
```

```{r summary of dat3}
summary(dat3)
```

## The 'mechanism' model

```{r mechanism model}
mech_model <- lme4::lmer(log_wesa ~ n_s * scale(flow) + dos + I(dos^2) + (dos + I(dos^2) | year),
                         data = dat3)
summary(mech_model)
```

### Diagnostic plots

```{r make mech plot data}
mech_plot <- broom.mixed::augment(mech_model, dat3)
mech_plot <- janitor::clean_names(mech_plot)
```

```{r mech plot: observed vs predicted}
ggplot(mech_plot, aes(x = log_wesa, y = fitted)) +
  geom_point() + 
  ggtitle("Observed vs. Fitted values") +
  xlab("Observed") + 
  ylab("Predicted") +
  theme_minimal()
```


```{r mech plot: heteroskedasticity}
ggplot(mech_plot, aes(x = fitted, y = resid)) +
  geom_point() + 
  ggtitle("Heteroskedasticity",
          subtitle = "Fitted values vs. Residuals") +
  xlab("Fitted") + 
  ylab("Residuals") +
  theme_minimal()
```


```{r mech plot: qq plot}
ggplot(mech_plot, aes(sample = resid)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Quantile-Quantile") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```


## The 'temporal' model

```{r temporal model}
temp_model <- lme4::lmer(log_wesa ~ n_s * year_c + dos + I(dos^2) + (dos + I(dos^2) | year),
                         data = dat3)
summary(temp_model)
```

### Diagnostic plots

```{r make temp plot data}
temp_plot <- broom.mixed::augment(temp_model, dat3)
temp_plot <- janitor::clean_names(temp_plot)
```

```{r temp plot: observed vs predicted}
ggplot(temp_plot, aes(x = log_wesa, y = fitted)) +
  geom_point() + 
  ggtitle("Observed vs. Fitted values") +
  xlab("Observed") + 
  ylab("Predicted") +
  theme_minimal()
```


```{r temp plot: heteroskedasticity}
ggplot(temp_plot, aes(x = fitted, y = resid)) +
  geom_point() + 
  ggtitle("Heteroskedasticity",
          subtitle = "Fitted values vs. Residuals") +
  xlab("Fitted") + 
  ylab("Residuals") +
  theme_minimal()
```


```{r temp plot: qq plot}
ggplot(temp_plot, aes(sample = resid)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Quantile-Quantile") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```

## Comparing the two

```{r}
anova(mech_model, temp_model)
```