---
title: "Brunswick Point Peep Models"
author: "Sarah Popov"
date: '`r Sys.Date()`'
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE)

library(ggplot2)

# Color palette (using blue & yellow from wesanderson Zissou1 palette)
pal <- c(rgb(86, 152, 175, maxColorValue = 255), rgb(229, 205, 79, maxColorValue = 255))

# Source the data file, which connects to the bppeeps database and then queries and processes the data
# This will produce two datasets: 
# 1) dat [counts + environmental covariates]
# 2) sr [daily species ratios]
source(here::here("models", "prepare_dat.R"))
```

\newpage
Roberts Bank, Delta, BC, is situated in the great Pacific Flyway and serves as an important stopover for peeps migrating north in the spring. It therefore hosts a large seasonal population of peeps: namely, Western sandpiper (WESA) and Dunlin (DUNL), that rely on the seasonal nutritional bounty provided by the Fraser River delta.

This document describes the suite of models used to estimate yearly changes in spatial distribution and abundance of peeps in the Roberts Bank estuary.

There are two datasets used in this modelling pipeline:

1. `sr`, which contains species ratio data (WESA:DUNL)
2. `dat`, which contains bird counts + environmental covariates

For an interactive interface with the `dat` dataset used in this document, see https://popovs.shinyapps.io/peepr/.

```{r sr binomial, child = "binomial_sr.Rmd"}
```

\newpage 

# Model data summary

The primary goal of this analysis is to determine whether melt flow regimes alter the north-south distribution of peeps. Birds are surveyed from a route of standardized survey stations. Any birds counted in or north of the "view corner" are labelled as `N`, while any birds counted south of the "view corner" are labelled as `S` for this analysis.^[For now, this includes birds counted in Canoe Pass, but those will be excluded in a later analysis to compare results.]

The `dat` dataset, which contains survey data and associated environmental covariates, will be used in this model. The `dat` dataset can be explored in greater detail in the [PeepR Shiny app](https://popovs.shinyapps.io/peepr/).

\newpage
### `dat` filtering

The `dat` dataset underwent some filtering steps to exclude unwanted data prior to this analysis. First, only data that was originally included in Canham et al. (2021) is extracted from the `bppeeps` database. While survey data oftentimes includes multiple sweeps, low-quality or reconnaissance sweeps were excluded from the daily totals used in Canham et al. (2021). As such, they were excluded from this dataset as well.

The total bird count for several surveys were obtained by taking the average total count from multiple sweeps. In these cases, it is difficult to obtain accurate location-level numbers because birds often moved locations between sweeps. These records are excluded for now (45 survey dates out of a total of 538).

Additionally, certain survey dates only included a total bird count for the day, but no location-level information. These records were excluded from the initial database query (24 survey dates out of a total of 538).

After excluding these surveys, the initial dataset queried from the `bppeeps` dataset included `r filtering$n_records[1]` records from a total of `r filtering$n_survey_dates[1]`. This is termed the 'full dataset' and was filtered further in `R`. The full dataset includes one survey per station per day. 

```{r filtering table}
knitr::kable(filtering, col.names = c("Filtering step", "No. survey dates", "No. records", "No. survey dates lost", "No. records lost"))
```

\newpage
## `dat` data summary after filtering

```{r dat summary}
summary(dat)
```

\newpage
```{r canham extended, child = "canham_extended.Rmd"}
```

\newpage
```{r mech vs time, child = "mech_vs_time.Rmd"}
```

\newpage

# WESA variable plots

```{r wesa vs year}
ggplot(dat3, aes(x = year_n, y = log_wesa, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Year vs log(WESA)",
       x = "Year", 
       y = "log(WESA)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

```{r wesa vs dos}
ggplot(dat3, aes(x = dos, y = log_wesa, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Day of Season vs log(WESA)",
       x = "DOS", 
       y = "log(WESA)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

```{r wesa vs flow}
ggplot(dat3, aes(x = flow, y = log_wesa, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Flow vs log(WESA)",
       x = "Flow", 
       y = "log(WESA)",
       color = "Station") + 
  cowplot::theme_cowplot()
```
```{r wesa vs u}
ggplot(dat3, aes(x = u, y = log_wesa, color = n_s)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "u (westerly winds) vs log(WESA)",
       x = "u (westerly winds)", 
       y = "log(WESA)",
       color = "Station") + 
  cowplot::theme_cowplot()
```
```{r wesa vs v}
ggplot(dat3, aes(x = v, y = log_wesa, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "v (southerly winds) vs log(WESA)",
       x = "v (southerly winds)", 
       y = "log(WESA)",
       color = "Station") + 
  cowplot::theme_cowplot()
```
```{r wesa vs windspd}
ggplot(dat3, aes(x = windspd, y = log_wesa, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Wind speed vs log(WESA)",
       x = "Wind speed (km/h)", 
       y = "log(WESA)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

```{r wesa vs tide}
ggplot(dat3, aes(x = elev_range, y = log_wesa, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Tidal amplitude vs log(WESA)",
       x = "Tidal amplitude (m)", 
       y = "log(WESA)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

\newpage

# WESA model

Full model:

```{r full wesa model}
full_wesa_model <- lmerTest::lmer("log_wesa ~ n_s*scale(flow) + n_s*scale(elev_range) + n_s*scale(u) + n_s*scale(v) + n_s*scale(windspd) + dos + I(dos^2) + year_c + (dos + I(dos^2) | year)",
           data = dat3)

summary(full_wesa_model)
```

Backwards stepwise selection:

```{r wesa stepwise}
lmerTest::step(full_wesa_model, data = dat3)
```

```{r final wesa model}
s <- lmerTest::step(full_wesa_model, data = dat3)
final_wesa_model <- lme4::lmer(lmerTest::get_model(s)@call$formula, data = dat3)
```


## Final model diagnostics

```{r wesa make temp plot data}
temp_plot <- broom.mixed::augment(final_wesa_model, dat3)
temp_plot <- janitor::clean_names(temp_plot)
```

```{r wesa temp plot: observed vs predicted}
ggplot(temp_plot, aes(x = log_wesa, y = fitted)) +
  geom_point() + 
  ggtitle("Observed vs. Fitted values") +
  xlab("Observed") + 
  ylab("Predicted") +
  theme_minimal()
```


```{r wesa temp plot: heteroskedasticity}
ggplot(temp_plot, aes(x = fitted, y = resid)) +
  geom_point() + 
  ggtitle("Heteroskedasticity",
          subtitle = "Fitted values vs. Residuals") +
  xlab("Fitted") + 
  ylab("Residuals") +
  theme_minimal()
```


```{r wesa temp plot: qq plot}
ggplot(temp_plot, aes(sample = resid)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Quantile-Quantile") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```

```{r wesa vars p1}
temp_plot %>%
  dplyr::select(year:resid, -fitted) %>%
  tidyr::gather(-resid, key = "var", value = "value") %>%
  dplyr::mutate(value = as.numeric(value)) %>%
  ggplot(aes(x = value, y = resid)) +
  geom_point() +
  scale_x_continuous(n.breaks = 3) +
  ggtitle("Variables vs. Residuals") + 
  ggforce::facet_wrap_paginate(~ var, ncol = 3, nrow = 3, scales = "free", page = 1) +
  theme_minimal()
```

```{r wesa vars p2}
temp_plot %>%
  dplyr::select(year:resid, -fitted) %>%
  tidyr::gather(-resid, key = "var", value = "value") %>%
  dplyr::mutate(value = as.numeric(value)) %>%
  ggplot(aes(x = value, y = resid)) +
  geom_point() +
  scale_x_continuous(n.breaks = 3) +
  ggforce::facet_wrap_paginate(~ var, ncol = 3, nrow = 3, scales = "free", page = 2) +
  theme_minimal()
```

```{r wesa vars p3}
temp_plot %>%
  dplyr::select(year:resid, -fitted) %>%
  tidyr::gather(-resid, key = "var", value = "value") %>%
  dplyr::mutate(value = as.numeric(value)) %>%
  ggplot(aes(x = value, y = resid)) +
  geom_point() +
  scale_x_continuous(n.breaks = 3) +
  ggforce::facet_wrap_paginate(~ var, ncol = 3, nrow = 3, scales = "free", page = 3) +
  theme_minimal()
```

# DUNL variable plots

```{r dunl vs year}
ggplot(dat3, aes(x = year_n, y = log_dunl, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Year vs log(DUNL)",
       x = "Year", 
       y = "log(DUNL)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

```{r dunl vs dos}
ggplot(dat3, aes(x = dos, y = log_dunl, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Day of Season vs log(DUNL)",
       x = "DOS", 
       y = "log(DUNL)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

```{r dunl vs flow}
ggplot(dat3, aes(x = flow, y = log_dunl, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Flow vs log(DUNL)",
       x = "Flow", 
       y = "log(DUNL)",
       color = "Station") + 
  cowplot::theme_cowplot()
```
```{r dunl vs u}
ggplot(dat3, aes(x = u, y = log_dunl, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "u (westerly winds) vs log(DUNL)",
       x = "u (westerly winds)", 
       y = "log(DUNL)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

```{r dunl vs v}
ggplot(dat3, aes(x = v, y = log_dunl, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "v (southerly winds) vs log(DUNL)",
       x = "v (southerly winds)", 
       y = "log(DUNL)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

```{r dunl vs windspd}
ggplot(dat3, aes(x = windspd, y = log_dunl, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Wind speed vs log(DUNL)",
       x = "Wind speed (km/h)", 
       y = "log(DUNL)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

```{r dunl vs tide}
ggplot(dat3, aes(x = elev_range, y = log_dunl, color = n_s)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Tidal amplitude vs log(DUNL)",
       x = "Tidal amplitude (m)", 
       y = "log(DUNL)",
       color = "Station") + 
  cowplot::theme_cowplot()
```

\newpage

# DUNL model

Full model:

```{r full dunl model}
full_dunl_model <- lmerTest::lmer("log_dunl ~ n_s*scale(flow) + n_s*scale(elev_range) + n_s*scale(u) + n_s*scale(v) + n_s*scale(windspd) + dos + I(dos^2) + year_c + (dos + I(dos^2) | year)",
           data = dat3)

summary(full_dunl_model)
```

Backwards stepwise selection:

```{r dunl stepwise}
lmerTest::step(full_dunl_model, data = dat3)
```

```{r final dunl model}
s <- lmerTest::step(full_dunl_model, data = dat3)
final_dunl_model <- lme4::lmer(lmerTest::get_model(s)@call$formula, data = dat3)
```

## Final model diagnostics

```{r dunl make temp plot data}
temp_plot <- broom.mixed::augment(final_dunl_model, dat3)
temp_plot <- janitor::clean_names(temp_plot)
```

```{r dunl temp plot: observed vs predicted}
ggplot(temp_plot, aes(x = log_wesa, y = fitted)) +
  geom_point() + 
  ggtitle("Observed vs. Fitted values") +
  xlab("Observed") + 
  ylab("Predicted") +
  theme_minimal()
```


```{r dunl temp plot: heteroskedasticity}
ggplot(temp_plot, aes(x = fitted, y = resid)) +
  geom_point() + 
  ggtitle("Heteroskedasticity",
          subtitle = "Fitted values vs. Residuals") +
  xlab("Fitted") + 
  ylab("Residuals") +
  theme_minimal()
```


```{r dunl temp plot: qq plot}
ggplot(temp_plot, aes(sample = resid)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Quantile-Quantile") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```

```{r dunl vars p1}
temp_plot %>%
  dplyr::select(year:resid, -fitted) %>%
  tidyr::gather(-resid, key = "var", value = "value") %>%
  dplyr::mutate(value = as.numeric(value)) %>%
  ggplot(aes(x = value, y = resid)) +
  geom_point() +
  scale_x_continuous(n.breaks = 3) +
  ggtitle("Variables vs. Residuals") + 
  ggforce::facet_wrap_paginate(~ var, ncol = 3, nrow = 3, scales = "free", page = 1) +
  theme_minimal()
```

```{r dunl vars p2}
temp_plot %>%
  dplyr::select(year:resid, -fitted) %>%
  tidyr::gather(-resid, key = "var", value = "value") %>%
  dplyr::mutate(value = as.numeric(value)) %>%
  ggplot(aes(x = value, y = resid)) +
  geom_point() +
  scale_x_continuous(n.breaks = 3) +
  ggforce::facet_wrap_paginate(~ var, ncol = 3, nrow = 3, scales = "free", page = 2) +
  theme_minimal()
```

```{r dunl vars p3}
temp_plot %>%
  dplyr::select(year:resid, -fitted) %>%
  tidyr::gather(-resid, key = "var", value = "value") %>%
  dplyr::mutate(value = as.numeric(value)) %>%
  ggplot(aes(x = value, y = resid)) +
  geom_point() +
  scale_x_continuous(n.breaks = 3) +
  ggforce::facet_wrap_paginate(~ var, ncol = 3, nrow = 3, scales = "free", page = 3) +
  theme_minimal()
```

\newpage
```{r yearly trend, child = "yearly_trend.Rmd"}
```