---
title: "Brunswick Point Peep Models"
author: "Sarah Popov"
date: "2022-11-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(ggplot2)

# Source the data file, which connects to the bppeeps database and then queries and processes the data
# This will produce two datasets: 
# 1) dat [counts + environmental covariates]
# 2) sr [daily species ratios]
source(here::here("models", "prepare_dat.R"))
```

Roberts Bank, Delta, BC, is situated in the great Pacific Flyway and serves as an important stopover for peeps migrating north in the spring. It therefore hosts a large seasonal population of peeps: namely, Western sandpiper (WESA) and Dunlin (DUNL), that rely on the seasonal nutritional bounty provided by the Fraser River delta.

This document describes the suite of models used to estimate yearly changes in spatial distribution and abundance of peeps in the Roberts Bank estuary.

There are two datasets used in this modelling pipeline:

1. `sr`, which contains species ratio data (WESA:DUNL)
2. `dat`, which contains bird counts + environmental covariates

For an interactive interface with the `dat` dataset used in this document, see https://popovs.shinyapps.io/peepr/.

### `sr` summary

```{r sr summary}
summary(sr)
```

\newpage
### `dat` summary

```{r dat summary}
summary(dat)
```

# Species composition model

The daily ratio of Western sandpiper (WESA) to Dunlin (DUNL) across the entire study period is first modelled using a dataset of known species ratios (species ratios are not measured during every survey).

The ratios are modelled using a binomial generalized linear mixed model (binomial GLMM). The resulting predicted ratios are then used to estimate the number of WESA vs. DUNL per day.

## Five models are built and compared

Response variable: 
* `y` - WESA:DUNL ratio

Predictor variables: 
* `dos` - day of season (recentered/scaled Julian date)
* `year` - year of survey

```{r sr glmm fitting}
# First, create our response variable: 
# WESA:DUNL proportion from our daily_percent_ratio table
y <- cbind(sr$wesa, sr$dunl)

# And build our 5 models 
# Response variable: ratio of WESA to DUNL
# Predictor variables:
# - dos: day of season (recentered/scaled julian date)
# - year: year (as a factor)
lme1 <- lme4::glmer(y ~ dos + I(dos^2) + (dos + I(dos^2)|year),
                    family = binomial, 
                    data = sr)
lme2 <- lme4::glmer(y  ~ dos + I(dos^2) + (1|year), 
                    family = binomial, 
                    data = sr)
lme3 <- lme4::glmer(y ~ dos + (dos|year), 
                    family = binomial, 
                    data = sr)
lme4 <- lme4::glmer(y ~ dos + (1|year), 
                    family = binomial, 
                    data = sr)
lme5 <- lme4::glmer(y ~ 1 + (1|year), 
                    family = binomial, 
                    data = sr)

## Compare AIC values 
anova(lme1, lme2, lme3, lme4, lme5)
```

The best-fit model is `lme1`. The residuals from `lme1` are appended to the `sr` dataset and another model is re-fit in order to estimate overdispersion. Because the standard deviation of the residuals is < 1, the model is deemed an appropriate candidate for predicting daily species ratios.

```{r best fit sr}
sr$resids <- residuals(lme1)

lme4::glmer(y ~ dos + I(dos^2) + (dos + I(dos^2)|year) + (1|resids),
            family = binomial, 
            data = sr)

sr_glmm <- lme1

# Add predicted values to sr
sr$predicted_ratio <- fitted(sr_glmm)
```

\newpage
### Summary of best fit model (`lme1`)

```{r summary best fit sr}
summary(sr_glmm)
```

\newpage
### Check assumptions of best fit model


```{r sr plot: observed vs predicted}
ggplot(sr, aes(x = p_wesa, y = predicted_ratio)) +
  geom_point() + 
  ggtitle("Observed vs. Predicted values") +
  xlab("Observed") + 
  ylab("Predicted") +
  theme_minimal()
```


```{r sr plot: heteroskedasticity}
ggplot(sr, aes(x = predicted_ratio, y = resids)) +
  geom_point() + 
  ggtitle("Heteroskedasticity",
          subtitle = "Fitted values vs. Residuals") +
  xlab("Fitted") + 
  ylab("Residuals") +
  theme_minimal()
```


```{r sr plot: qq plot}
ggplot(sr, aes(sample = resids)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Quantile-Quantile") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```


```{r sr year effects, include = FALSE}
year_effects <- lme4::ranef(sr_glmm)$year
year_effects <- janitor::clean_names(year_effects)
```


```{r sr plot: year intercept plot}
ggplot(year_effects, aes(sample = intercept)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Year effects", 
          subtitle = "Intercept") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```


```{r sr plot: year slope plot}
ggplot(year_effects, aes(sample = dos)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Year effects", 
          subtitle = "Slope") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```


```{r sr plot: year slope quadratic plot}
ggplot(year_effects, aes(sample = i_dos_2)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Year effects", 
          subtitle = "Slope quadratic") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```

\newpage
## Predict WESA/DUNL population

Using the derived binomial GLMM above, we will predict the amount of WESA and DUNL each day. For any years that are missing from the bGLMM we will assign the mean proportion of WESA:DUNL. 

```{r predict wesa ratio, include = FALSE}
# Create yrs df, which is a df containing all possible
# julian_day-year combinations
yrs <- expand.grid(julian_day = seq(min(dat$julian_day), max(dat$julian_day), 1),
                   year = as.factor(unique(lubridate::year(dat$survey_date))))
# Add day of season variable
yrs$dos <- (yrs$julian_day - mean(sr$julian_day))/sd(sr$julian_day)

# Extract sr_glmm model coefficients
coefs <- coef(sr_glmm)$year
names(coefs) <- c("intercept", "jd_parm", "jd2_parm")
coefs$year <- unique(yrs$year)

# Merge yrs and coefs together
yrs <- merge(yrs, coefs, by = "year", all.x = TRUE)

# Previously, within Canham paper, the mean coefficients across all
# years were calculated for years without species ratios. However,
# in this case that was not necessary, as there are no missing years
# in the species ratio/count data.

# Now calculate odds & predict ratios for all survey dates
yrs$odds <- yrs$intercept + yrs$jd_parm * yrs$dos + yrs$jd2_parm * yrs$dos^2
yrs$predicted_ratio <- exp(yrs$odds) / (1 + exp(yrs$odds))
```

```{r plot predicted wesa ratios, fig.height = 8}
ggplot() +
  geom_line(data = yrs, aes(x = julian_day, y = predicted_ratio, group = year)) +
  geom_point(data = sr, aes(x = julian_day, y = (p_wesa/100), group = year), color = 'blue') +
  facet_wrap(~ year, ncol = 3)
```