---
title: "NS binomial"
author: "Sarah Popov"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE)

library(ggplot2)

# Source the data file, which connects to the bppeeps database and then queries and processes the data
# This will produce two datasets: 
# 1) dat [counts + environmental covariates]
# 2) sr [daily species ratios]
source(here::here("models", "prepare_dat.R"))

## GLMM N-S model
# First, create our dataset
ns_dat <- tidyr::pivot_wider(dat[,c("year", "julian_day", "survey_date", "n_s", "final_count", "elev_range", "flow", "total_precip", "mean_temp", "u", "v")], 
                   id_cols = c("year", "julian_day", "survey_date"),
                   names_from = "n_s",
                   values_from = "final_count",
                   values_fn = sum,
                   unused_fn = mean)

# Remove NA or 0 in N/S
ns_dat <- ns_dat[!is.na(ns_dat$N),]
ns_dat <- ns_dat[!is.na(ns_dat$S),]
ns_dat <- ns_dat[which(ns_dat$N != 0),]
ns_dat <- ns_dat[which(ns_dat$S != 0),]

# Rescale/center some vars
ns_dat$dos <- scale(ns_dat$julian_day)
ns_dat$log_N <- log(ns_dat$N + 1)
ns_dat$log_S <- log(ns_dat$S + 1)

# Extract complete cases of ns_dat
ns_dat <- ns_dat[complete.cases(ns_dat),]
```

Testing out a binomial GLMM where the response variable is total number of birds in North vs. South stations.

```{r summary ns_dat}
summary(ns_dat)
```

Check for correlation in response variables

```{r res cor}
# Correlation matrix in response ('res') variables
res <- ns_dat[,c("elev_range", "flow", "total_precip", "mean_temp", "u", "v")]
cor(res, use = "complete.obs")
```

Simple model with day of season and year - same exact models as in the WESA ratio model

```{r 5 models}
# Now, create our response variable: 
# N:S ratio - at a given survey time, N birds at northern vs southern station
ns <- cbind(ns_dat$N, ns_dat$S)

# And build our 5 models 
# Response variable: ratio of count of N birds to count of S birds
# Predictor variables:
# - dos: day of season (recentered/scaled julian date)
# - year: year (as a factor)
lme1 <- lme4::glmer(ns ~ dos + I(dos^2) + (dos + I(dos^2)|year),
                    family = binomial, 
                    data = ns_dat)
lme2 <- lme4::glmer(ns  ~ dos + I(dos^2) + (1|year), 
                    family = binomial, 
                    data = ns_dat)
lme3 <- lme4::glmer(ns ~ dos + (dos|year), 
                    family = binomial, 
                    data = ns_dat)
lme4 <- lme4::glmer(ns ~ dos + (1|year), 
                    family = binomial, 
                    data = ns_dat)
lme5 <- lme4::glmer(ns ~ 1 + (1|year), 
                    family = binomial, 
                    data = ns_dat)

## Compare AIC values 
anova(lme1, lme2, lme3, lme4, lme5)
```

Overall the models are quite similar. In this case I simply just chose `lme1` as a base model to compare the effects of environmental variables, which are below. All models except for `flow_glmm` result in a `Model is nearly unidentifiable: very large eigenvalue. Rescale variables?` warning.

```{r lme1 + env variables}
flow_glmm <- lme4::glmer(ns ~ dos + I(dos^2) + (dos + I(dos^2)|year) + scale(flow),
                    family = binomial, 
                    data = ns_dat,
                    control = lme4::glmerControl(optimizer = "bobyqa"))

elev_glmm <- lme4::glmer(ns ~ dos + I(dos^2) + (dos + I(dos^2)|year) + scale(elev_range),
                         family = binomial, 
                         data = ns_dat,
                         control = lme4::glmerControl(optimizer = "bobyqa"))

precip_glmm <- lme4::glmer(ns ~ dos + I(dos^2) + (dos + I(dos^2)|year) + scale(total_precip),
                         family = binomial, 
                         data = ns_dat,
                         control = lme4::glmerControl(optimizer = "bobyqa"))

temp_glmm <- lme4::glmer(ns ~ dos + I(dos^2) + (dos + I(dos^2)|year) + scale(mean_temp),
                           family = binomial, 
                           data = ns_dat,
                           control = lme4::glmerControl(optimizer = "bobyqa"))

u_glmm <- lme4::glmer(ns ~ dos + I(dos^2) + (dos + I(dos^2)|year) + scale(u),
                         family = binomial, 
                         data = ns_dat,
                         control = lme4::glmerControl(optimizer = "bobyqa"))

v_glmm <- lme4::glmer(ns ~ dos + I(dos^2) + (dos + I(dos^2)|year) + scale(v),
                      family = binomial, 
                      data = ns_dat,
                      control = lme4::glmerControl(optimizer = "bobyqa"))

## Compare AIC values 
anova(lme1, flow_glmm, elev_glmm, precip_glmm, temp_glmm, u_glmm, v_glmm)
```

\newpage

## Flow GLMM summary

```{r flow glmm summary}
summary(flow_glmm)
```

\newpage
## Flow GLMM fit plots

```{r add resids and predicted to ns_dat}
ns_dat$resids <- resid(flow_glmm)
ns_dat$ns_ratio <- ns_dat$N / (ns_dat$N + ns_dat$S)
ns_dat$predicted_ratio <- fitted(flow_glmm)
```

```{r plot: observed vs predicted}
ggplot(ns_dat, aes(x = ns_ratio, y = predicted_ratio)) +
  geom_point() + 
  ggtitle("Observed vs. Predicted values") +
  xlab("Observed") + 
  ylab("Predicted") +
  theme_minimal()
```

```{r plot: heteroskedasicity}
ggplot(ns_dat, aes(x = predicted_ratio, y = resids)) +
  geom_point() + 
  ggtitle("Heteroskedasticity",
          subtitle = "Fitted values vs. Residuals") +
  xlab("Fitted") + 
  ylab("Residuals") +
  theme_minimal()
```

```{r qqplot}
ggplot(ns_dat, aes(sample = resids)) +
  stat_qq() + 
  stat_qq_line() +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  ggtitle("Quantile-Quantile") +
  xlab("Theoretical") + 
  ylab("Sample") +
  theme_minimal()
```

## Flow GLMM residuals vs. other predictors

```{r plot: resid vs elev_range}
ggplot(ns_dat, aes(y = resids, x = elev_range)) +
  geom_point() + 
  ggtitle("Tidal amplitude vs Residuals") +
  ylab("Residuals") + 
  xlab("Tidal amplitude") +
  theme_minimal()
```

```{r plot: resid vs precip}
ggplot(ns_dat, aes(y = resids, x = total_precip)) +
  geom_point() + 
  ggtitle("Total precipitation vs Residuals") +
  ylab("Residuals") + 
  xlab("Total precipitation") +
  theme_minimal()
```

```{r plot: resid vs mean_temp}
ggplot(ns_dat, aes(y = resids, x = mean_temp)) +
  geom_point() + 
  ggtitle("Mean temp vs Residuals") +
  ylab("Residuals") + 
  xlab("Mean temp") +
  theme_minimal()
```

```{r plot: resid vs u}
ggplot(ns_dat, aes(y = resids, x = u)) +
  geom_point() + 
  ggtitle("u vs Residuals") +
  ylab("Residuals") + 
  xlab("u") +
  theme_minimal()
```

```{r plot: resid vs v}
ggplot(ns_dat, aes(y = resids, x = v)) +
  geom_point() + 
  ggtitle("v vs Residuals") +
  ylab("Residuals") + 
  xlab("v") +
  theme_minimal()
```