---
title: "Tweedie distribution"
author: "Sarah Popov"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup tweedie, include=FALSE}
library(ggplot2)
library(formatR)

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE,
                      tidy.opts = list(width.cutoff = 60), 
                      tidy = TRUE)

# Color palette (using blue & yellow from wesanderson Zissou1 palette)
pal <- c(rgb(86, 152, 175, maxColorValue = 255), rgb(229, 205, 79, maxColorValue = 255))

# Source the data file, which connects to the bppeeps database and then queries and processes the data
# This will produce two datasets: 
# 1) dat [counts + environmental covariates]
# 2) sr [daily species ratios]
source(here::here("models", "prepare_dat.R"))
source(knitr::purl(here::here("models", "child_docs", "binomial_sr.Rmd"), quiet = TRUE))
source(here::here("models", "prepare_dat_ns.R"))
```


```{r p0, include = FALSE}
# Proportion of zeroes
dat_p0 <- sum(dat$predicted_wesa == 0) / length(dat$predicted_wesa)
dat_ns_p0 <- sum(dat_ns$predicted_wesa == 0) / length(dat_ns$predicted_wesa)
```

```{r wesa gam base models, include=TRUE, echo=TRUE}
# Base script by Gavin Simpson
# https://fromthebottomoftheheap.net/2017/05/04/compare-mgcv-with-glmmtmb/
# https://gist.github.com/gavinsimpson/8a0f0e072b095295cf5f7af2762e05a7

library("mgcv")
library("glmmTMB")

## Zero-inflated negative binomial
# This was our set of the previous best-fit models
zinb0 <- glmmTMB(predicted_wesa ~ n_s + year_c + I(dos^2) + (1|year), zi = ~ n_s, data = dat_ns, family = nbinom1)
zinb1 <- glmmTMB(predicted_wesa ~ n_s + scale(flow) + year_c + I(dos^2) + (1|year), zi = ~ n_s + scale(flow), data = dat_ns, family = nbinom1)
zinb2 <- glmmTMB(predicted_wesa ~ n_s * scale(flow) + year_c + I(dos^2) + (1|year), zi = ~ n_s * scale(flow), data = dat_ns, family = nbinom1)

AIC(zinb0, zinb1, zinb2)

## Tweedie GAM
tgam0 <- gam(predicted_wesa ~ n_s + s(year_c) + s(dos) + s(year, bs = "re"), data = dat_ns, family = tw, method = "ML")
tgam1 <- gam(predicted_wesa ~ n_s + s(flow) + s(year_c) + s(dos) + s(year, bs = "re"), data = dat_ns, family = tw, method = "ML")
tgam2 <- gam(predicted_wesa ~ n_s + s(flow, by = n_s) + s(year_c) + s(dos) + s(year, bs = "re"), data = dat_ns, family = tw, method = "ML")

AIC(tgam0, tgam1, tgam2)

## Zero-inflated tweedie GLM
zit0 <- glmmTMB(predicted_wesa ~ n_s + year_c + I(dos^2) + (1|year), zi = ~ n_s, data = dat_ns, family = tweedie)
zit1 <- glmmTMB(predicted_wesa ~ n_s + scale(flow) + year_c + I(dos^2) + (1|year), zi = ~ n_s + scale(flow), data = dat_ns, family = tweedie)
zit2 <- glmmTMB(predicted_wesa ~ n_s * scale(flow) + year_c + I(dos^2) + (1|year), zi = ~ n_s * scale(flow), data = dat_ns, family = tweedie)

AIC(zit0, zit1, zit2)

# Compare them all
bbmle::AICtab(zinb0, zinb1, zinb2, tgam0, tgam1, tgam2, zit0, zit1, zit2)
```

## Best-fit diagnostics

```{r}
DHARMa::testResiduals(tgam2, plot = T)
```

Test for zero inflation

```{r, fig.cap="The zero-inflation test suggests the Tweedie distribution is doing a good job predicting zeroes."}
resid_sim <- DHARMa::simulateResiduals(tgam2)
DHARMa::testZeroInflation(resid_sim)
```

## Full model

```{r full model}
full <- gam(predicted_wesa ~ n_s + s(flow) + s(flow, by = n_s) + s(year_c) + s(dos)
            + s(mean_temp) + s(mean_temp, by = n_s) + s(elev_range) + s(elev_range, by = n_s) 
            + s(total_precip) + s(total_precip, by = n_s) + s(u) + s(u, by = n_s) 
            + s(windspd) + s(windspd, by = n_s) + tide + s(year, bs = "re"), 
                     data = dat_ns, 
                     family = tw)
summary(full)
```


```{r model selection, include = FALSE}
build <- buildmer::buildgam(predicted_wesa ~ n_s + s(flow, by = n_s) + s(year_c) 
                            + s(dos) + s(mean_temp) + s(elev_range) 
                            + s(total_precip) + s(u) + s(windspd) + tide 
                            + s(year, bs = "re"), 
                            dat_ns, 
                            family = tw, 
                            buildmerControl = list(direction = 'order'))

(f <- formula(build@model))

m <- buildmer::buildgam(f, 
                        data = dat_ns, 
                        buildmerControl = list(direction = 'backward'))

final_f <- formula(m@model)

build_mod <- gam(f, data = dat_ns, family = tw)
build_mod_f <- gam(final_f, data = dat_ns, family = tw)

bbmle::AICtab(full, build_mod, build_mod_f)
```

## Final model

Backwards stepwise selection; first removed insignificant terms from zi model, then subsequently removed insignificant terms from full model using AIC backwards selection (`drop1` command).

```{r final model summary}
summary(final_mod)
```

```{r final model plot}
sjPlot::plot_model(final_mod)
```

Do the zeroes in our ZI model make sense? At least for the year and N/S components, it seems so. In earlier years, the odds of a zero-bird count are higher, and in the south, the odds of a zero-bird count are also higher. With a zero-inflated model these odds are specifically the odds of finding *structural* or *true* zeroes, rather than simply zeroes due to sampling errors.

```{r, fig.cap="Frequency of zero count observations by year and north/south region. Zero counts are more frequently observed in earlier years and in the south."}
table(dat_ns[["year"]][dat_ns$final_count == 0], dat_ns[["n_s"]][dat_ns$final_count == 0], dat_ns[["final_count"]][dat_ns$final_count == 0])
```


## Final model diagnostics

```{r diagnostics, fig.cap="Residual diagnostics."}
DHARMa::testResiduals(final_mod)
```

```{r, fig.cap="Testing for overdispersion. Still not quite predicting the number of zeroes exactly correctly but better than before."}
resid_sim <- DHARMa::simulateResiduals(final_mod)
DHARMa::testZeroInflation(resid_sim)
```


### Residuals vs. predicted

```{r residuals vs predicted}
varsList <- c("dos", "year_c", "year", "n_s", "tide", "flow", "mean_temp", "elev_range", "total_precip", "u")
for (i in varsList) {
  cat("\n", i)
  DHARMa::plotResiduals(final_mod, dat_ns[[i]], rank = TRUE)
}
```

