---
title: "Yearly trends"
author: "Sarah Popov"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r yearly trends setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE)

# This RMD is NOT intended as a standalone script, but rather to be sourced
# from models.Rmd. The data used in this Rmd file is generated in models.Rmd
# and binomial_sr.Rmd.

# Run binomial sr model on the daily count data
# NOTE THE RESULTS WILL BE SLIGHTLY DIFFERENT TO THE CANHAM PAPER BECAUSE THIS 
# SR MODEL INCLUDES 2021-2022.

# Now merge predicted wesa ratio generated above to the dt dataset
dt <- merge(dt, yrs[,c("year", "julian_day", "predicted_ratio")], 
            by = c("year", "julian_day"))

# Scale and log-transform variables of interest
dt$predicted_wesa <- round(dt$total_count * dt$predicted_ratio, 0)
dt$predicted_dunl <- dt$total_count - dt$predicted_wesa
dt$log_wesa <- log(dt$predicted_wesa + 1)
dt$log_dunl <- log(dt$predicted_dunl + 1)
dt$year_n <- as.numeric(dt$year)
dt$year_c <- scale(as.numeric(dt$year))
```

# Yearly trends in peep population

The original Canham et al. (2021) paper provided yearly population trends of WESA and DUNL at Brunswick Point. These same population trends are extended here with 2020-2022 data, using the same population models derived in the original paper (i.e., no north vs. south component). Values represent population indices (with 95% confidence intervals) calculated as predicted values for each year from final models for each species, with independent variables held at median values for each year. Daily total counts (not broken down by survey station) are used as the data for this model.

*WESA model*:

$log(WESA) \sim year + dos + I(dos^2) + scale(elev.range) + scale(flow) + scale(u) + (dos + I(dos^2) | year)$

*DUNL model*:

$log(DUNL) \sim year + dos + I(dos^2) + scale(elev.range) + scale(flow) + scale(u) + (dos + I(dos^2) | year)$

```{r canham yearly trend model}
wesa_trend <- lmerTest::lmer(log_wesa ~ year_c + dos + I(dos^2) + scale(elev_range) + scale(flow) + scale(u) + (dos + I(dos^2) | year), 
                   data = dt,
                   REML = T, 
                   lme4::lmerControl(optimizer = 'bobyqa'))

dunl_trend <- lmerTest::lmer(log_dunl ~ year_c + dos + I(dos^2) + scale(elev_range) + scale(flow) + scale(u) + (dos + I(dos^2) | year), 
                   data = dt,
                   REML = T, 
                   lme4::lmerControl(optimizer = 'bobyqa')) 
```


```{r index df setup, include=FALSE}
# index0 - index df with same median values across *all* years
# - this will be fed into trend models for predictions
index0 <- data.frame(year = unique(dt$year),
                     elev_range = median(dt$elev_range, na.rm = T),
                     flow = median(dt$flow, na.rm = T),
                     u = median(dt$u, na.rm = T),
                     dos = median(dt$dos, na.rm = T))
index0$year_c <- (as.numeric(index0$year) - mean(dt$year_n)) / sd(dt$year_n) # equivalent to unique(dt$year_c)
index0$year_fac <- 1

# index - median covariates per year
index <- dt %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(dos = median(dos, na.rm = T),
                   elev_range = median(elev_range, na.rm = T),
                   flow = median(flow, na.rm = T),
                   u = median(u, na.rm = T),
                   v = median(v, na.rm = T),
                   windspd = median(windspd, na.rm = T)) %>%
  dplyr::mutate(year_n = as.numeric(as.character(year)),
                year_c = scale(year_n)) %>%
  as.data.frame()

# Calculate predicted WESA/DUNL numbers for each year
# using index0 values for the prediction
# NOTE this is specifically using merMod::predict
# param 're.form = NA' indicates to include NO random effects
index$pred_wesa <- exp(predict(wesa_trend, index0, re.form = NA))
index$pred_dunl <- exp(predict(dunl_trend, index0, re.form = NA))

# Calculate fixed-effects only se for WESA/DUNL numbers for each year, incl residual error
# using index0 for predictions
index[,c("wesa_fit_t", "wesa_upr_t", "wesa_lwr_t")] <- 
  exp(merTools::predictInterval(merMod = wesa_trend,
                                newdata = index0,
                                level = 0.95,
                                n.sims = 1000,
                                which = "fixed",
                                stat = "median",
                                type = "linear.prediction",
                                include.resid.var = FALSE)) / 1000

index[,c("dunl_fit_t", "dunl_upr_t", "dunl_lwr_t")] <- 
  exp(merTools::predictInterval(merMod = dunl_trend,
                                newdata = index0,
                                level = 0.95,
                                n.sims = 1000,
                                which = "fixed",
                                stat = "median",
                                type = "linear.prediction",
                                include.resid.var = FALSE)) / 1000

# Calculate year indices, incl random effect, but no residual error
# using index for predictions
index[,c("wesa_fit", "wesa_upr", "wesa_lwr")] <- 
  exp(merTools::predictInterval(merMod = wesa_trend, 
                                newdata = index,
                                level = 0.95, 
                                n.sims = 1000,
                                stat = "median", 
                                type="linear.prediction",
                                include.resid.var = FALSE)) / 1000

index[,c("dunl_fit", "dunl_upr", "dunl_lwr")] <- 
  exp(merTools::predictInterval(merMod = dunl_trend, 
                                newdata = index,
                                level = 0.95, 
                                n.sims = 1000,
                                stat = "median", 
                                type="linear.prediction",
                                include.resid.var = FALSE)) / 1000
```

```{r percent change, include=FALSE}
## define functions for estimates of percent change  
perc_change<-function(x,i){
      ((x[length(x)]-x[i])/abs(x[i]))*100
    }
    
perc_change.yr<-function(x,i){
      (((x[length(x)]-x[i])/abs(x[i]))*100)/length(x)
    }

## western sandpiper
perc_change(index$pred_wesa, 1)  # from start 	
perc_change.yr(index$pred_wesa, 1)
 
perc_change(index$pred_wesa, 20)  # from 2012 	
perc_change.yr(index$pred_wesa, 20) 
  
## dunlin 
perc_change(index$pred_dunl, 1)  # from start 	
perc_change.yr(index$pred_dunl, 1)
 
perc_change(index$pred_dunl, 20)  # from 2012 	
perc_change.yr(index$pred_dunl, 20) 
```

## WESA population trend

Western sandpiper populations continue to show yearly declines in population. Since 1991, western sandpiper populations have fallen by `r round(perc_change(index$pred_wesa, 1), 1) * -1`% (`r round(perc_change.yr(index$pred_wesa, 1), 1)`% per annum). In the last ten years the rate of decline has slowed. Since 2012, western sandpiper populations have fallen by `r round(perc_change(index$pred_wesa, 20), 1) * -1`% (`r round(perc_change.yr(index$pred_wesa, 20), 1)`% per annum).

```{r wesa trend model summary}
summary(wesa_trend)
```

```{r wesa index plot}
index %>%
  ggplot() +
  geom_ribbon(aes(x = year_n, 
                  y = wesa_fit_t,
                  ymin = wesa_lwr_t,
                  ymax = wesa_upr_t),
              color = NA,
              fill = 'lightgrey') + 
  geom_line(aes(x = year_n,
                y = wesa_fit_t),
            color = 'red',
            linewidth = 1) +
  geom_pointrange(aes(x = year_n, 
                      y = wesa_fit, 
                      ymin = wesa_lwr, 
                      ymax = wesa_upr),
                  fatten = 2) +
  scale_y_continuous(trans = 'log10') +
  labs(title = 'Western Sandpiper', 
       x = 'Year', 
       y = expression(paste("Index (", 10^3, " birds; ", log[10],  " scale)"))
       ) +
  cowplot::theme_cowplot()
```

## DUNL population trend

As in the original Canham paper, however, `year` remains insignificant in the DUNL model.

Pacific Dunlin populations continue to show yearly declines in population, though the trend is not statistically significant. Since 1991, Pacific Dunline populations have fallen by `r round(perc_change(index$pred_dunl, 1), 1) * -1`% (`r round(perc_change.yr(index$pred_dunl, 1), 1)`% per annum). In the last ten years the rate of decline has slowed. Since 2012, Pacific Dunline populations have fallen by `r round(perc_change(index$pred_dunl, 20), 1) * -1`% (`r round(perc_change.yr(index$pred_dunl, 20), 1)`% per annum).

```{r dunl trend model summary}
summary(dunl_trend)
```

```{r dunl index plot}
index %>%
  ggplot() +
  geom_ribbon(aes(x = year_n, 
                  y = dunl_fit_t,
                  ymin = dunl_lwr_t,
                  ymax = dunl_upr_t),
              color = NA,
              fill = 'lightgrey') + 
  geom_line(aes(x = year_n,
                y = dunl_fit_t),
            color = 'red',
            linewidth = 1,
            linetype = "dashed") +
  geom_pointrange(aes(x = year_n, 
                      y = dunl_fit, 
                      ymin = dunl_lwr, 
                      ymax = dunl_upr),
                  fatten = 2) +
  scale_y_continuous(trans = 'log10') +
  labs(title = 'Pacific Dunlin', 
       x = 'Year', 
       y = expression(paste("Index (", 10^3, " birds; ", log[10],  " scale)"))
       ) +
  cowplot::theme_cowplot()
```